### 2021-09-12 draft.tex
Deploying to gh-pages from @ PoisotLab/ms_metaweb_transfer@7e1fb30781c47bd0a86a8dcf9e9ef05c54084b9a 🚀

\usepackage{amsmath,amssymb,amsfonts}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{stix}
\usepackage[scaled]{helvet}
\usepackage[scaled]{inconsolata}

@@

\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[3] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
  % set entry spacing
  \ifnum #2 > 0
  \setlength{\parskip}{#2\baselineskip}
  \fi
 }%
 {}
\usepackage{calc} % for \widthof, \maxof
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\maxof{\widthof{#1}}{\csllabelwidth}}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth}{#1}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}\geometry{verbose,letterpaper,tmargin=2.5cm,bmargin=2.5cm,lmargin=2.5cm,rmargin=4.5cm}

@@

@@

\fancypagestyle{normal}
{
  \fancyhf{}
  \fancyfoot[R]{\footnotesize\sffamily\thepage\ of \pageref*{LastPage}}
}
\begin{document}
\thispagestyle{empty}
{\Large\bfseries\sffamily Food web reconstruction through phylogenetic
transfer of low-rank network representation}
\vskip 5em

%
\href{https://orcid.org/0000-0001-6067-1349}{Tanya\,Strydom}%
%
\,\textsuperscript{1,2,‡}\quad %
\href{https://orcid.org/0000-0003-0193-5441}{Salomé\,Bouskila}%
%
\,\textsuperscript{1,‡}\quad %
\href{https://orcid.org/0000-0001-9051-0597}{Francis\,Banville}%
%
\,\textsuperscript{1,3,2}\quad %
\href{https://orcid.org/0000-0003-4036-977X}{Ceres\,Barros}%
%
\,\textsuperscript{4}\quad %
\href{https://orcid.org/0000-0002-2151-6693}{Dominique\,Caron}%
%
\,\textsuperscript{5,2}\quad %
\href{https://orcid.org/0000-0003-0452-6993}{Maxwell J\,Farrell}%
%
\,\textsuperscript{6}\quad %
\href{https://orcid.org/0000-0002-9935-1366}{Marie-Josée\,Fortin}%
%
\,\textsuperscript{6}\quad %
\href{https://orcid.org/0000-0003-3220-6161}{Victoria\,Hemming}%
%
\,\textsuperscript{4}\quad %
\href{https://orcid.org/0000-0002-4104-9463}{Benjamin\,Mercier}%
%
\,\textsuperscript{3,2}\quad %
\href{https://orcid.org/0000-0002-6004-4027}{Laura J.\,Pollock}%
%
\,\textsuperscript{5}\quad %
\href{https://orcid.org/0000-0001-5785-8321}{Rogini\,Runghen}%
%
\,\textsuperscript{7}\quad %
\href{https://orcid.org/0000-0002-3454-0633}{Giulio V.\,Dalla Riva}%
%
\,\textsuperscript{8}\quad %
\href{https://orcid.org/0000-0002-0735-5184}{Timothée\,Poisot}%
%
\,\textsuperscript{1,2}

@@

\vfill
\textbf{\sffamily Abstract: }Despite their importance in many ecological
processes, collecting data and information on ecological interactions,
and therefore species interaction networks, is an exceedingly
challenging task. For this reason, large parts of the world have a
deficit of data of which species interact, and what we can expect the
network structure of these interactions to be. As data collection alone
is unlikely to be sufficient at filling these global gaps, community
ecologists must adopt predictive methods. In this contribution we
develop such a method, relying on graph embedding (the extraction of
explanatory latent variables from known graph structures) and transfer
learning (the application of previous solution to novel problems with
limited predictors overlap) in order to assemble a predicted list of
trophic interactions between mammals of Canada. This interaction list is
derived from extensive knowledge of the mammalian food web of Europe,
despite the fact that there are fewer than 5\% of common species between
the two locations. We provide guidance on how this method can be adapted
by substituting some approaches or predictors in order to make it more
generally applicable to a broad family of ecological problems.
\vfill

@@

There are two core challenges we are faced with in furthering our
understanding of ecological networks across space, particularly at
macro-ecologically relevant scales (\emph{e.g.} Trøjelsgaard and Olesen
2016). First, networks within a location are difficult to sample
properly (Jordano 2016a, 2016b), resulting in a widespread ``Eltonian
shortfall'' (Hortal et al. 2015). This first challenge (local
incompleteness) has been, in large part, addressed by the recent
multiplication of methods aiming to predict interactions \emph{within}
an \emph{existing} network, a lot of which are reviewed in Strydom et
al. (2021). Second, recent analyses based on collected data (Poisot,
Bergeron, et al. 2021) or metadata (Cameron et al. 2019) highlight that
ecological networks are currently studied in a biased subset of space
and bioclimates, which impedes our ability to generalize any local
understanding of network structure. Meaning that although the framework
to address incompleteness \emph{within} a network exists there would
still be regions that, due to a \emph{lack} of local interaction data,
we are unable to infer potential species interactions. Having a general
solution for the issue of metaweb inference (Morales-Castilla et al.
2015) that, despite situations where minimal knowledge about
interactions within a species pool is known, is capable of producing a
plausible metaweb could be the catalyst for significant breakthroughs in
our ability to start thinking about species interactions networks over
large spatial scales.

Here, we present a general method for the transfer learning of network
representations, relying on the similarities of species in a
biologically/ecologically relevant proxy space (\emph{e.g.} shared
morphology or ancestry). Transfer learning is a machine learning
methodology that uses the knowledge gained from solving one problem and
applying it to a related (destination) problem (Torrey and Shavlik 2010;
Pan and Yang 2010). In this instance, we solve the problem of predicting
trophic interactions between species, based on knowledge extracted from
another species pool for which interactions are known, using
phylogenetic structure as a medium for transfer. This allows us to
construct a \emph{probabilistic} metaweb for a community for which we
have \emph{no} prior interaction data for the desired species pool. Our
methodology is outlined in fig.~\ref{fig:concept}, where we provide an
illustration based on learning an embedding of a metaweb of trophic
interactions for European mammals (known interactions; Maiorano et al.
2020b, 2020a), and based on phylogenetic relationships between mammals
globally (Upham, Esselstyn, and Jetz 2019), infer a metaweb for the
Canadian mammalian species pool (interactions are treated as unknown in
this instance).

@@

There is a plurality of measures of species similarities that can be
used for metaweb reconstruction (see \emph{e.g.} Morales-Castilla et al.
2015); however, phylogenetic proximity has several desirable properties
when working at large scales. Gerhold et al. (2015) made the point that
phylogenetic signal captures diversification of characters (large
macro-evolutionary process), but not necessarily community assembly
(fine ecological process); Dormann et al. (2010) previously found very
similar conclusions. Interactions tend to conserve a phylogenetic signal
that encompasses a wide range of ecological and evolutionary mechanisms
(Mouquet et al. 2012; Cavender-Bares et al. 2009), and - most
importantly - retain this signal even when it is not detectable at the
community scale (Poisot and Stouffer 2018; Hutchinson, Cagua, and
Stouffer 2017). Finally, species interactions at macro-ecological scales
seem to respond mostly to macro-evolutionary processes (Price 2003);
which is evidenced by the presence of conserved backbones in food webs
(Dalla Riva and Stouffer 2016), strong evolutionary signature on prey
choice (Stouffer et al. 2012), and strong phylogenetic signature in food
web intervality (Eklöf and Stouffer 2016). Phylogenetic reconstruction
has also previously been used to understand ancestral plant-insect
interaction networks (Braga et al. 2021). Taken together, these
considerations suggest that phylogenies can reliably be used to transfer
knowledge on species interactions.

Our case study shows that phylogenetic transfer learning is indeed an
effective approach to predict the Canadian mammalian metaweb. This
showcases that although the components (species) that make up the
Canadian and European communities may not be \emph{perfectly} shared, if
the medium (proxy space) selected in the transfer step is biologically
plausible, we can still effectively learn from the known network and
make biologically relevant predictions of interactions. It should be
reiterated that the framework presented in fig.~\ref{fig:concept} is
amenable to changes; notably, the measure of similarity may not be
phylogeny, and can be replaced by information on foraging (Beckerman,
Petchey, and Warren 2006), cell-level mechanisms (Boeckaerts et al.
2021), or a combination of traits and phylogenetic structure (Stock
2021).

@@

We use data on the European metaweb assembled by Maiorano et al.
(2020b), following the definition of the metaweb first introduced by
Dunne (2006), \emph{i.e.} an inventory of all possible interactions
within species from a spatially delimited pool. Notably the metaweb is
not a prediction of the food web at any specific locale within the
frontiers of the species pool -- in fact, these local food webs are
expected to have a subset of both the species and the interactions of
their metaweb (Poisot et al. 2012). This being said, as the metaweb
represents the total of functional, phylogenetic, and macroecological
processes (Morales-Castilla et al. 2015), it is thus still worthy of
ecological attention. We induced the subgraph corresponding to all
mammals by matching species names in the original network first to the
GBIF taxonomic backbone (GBIF Secretariat 2021) and retaining all those
who matched to mammals; all nodes had valid matches to GBIF at this
step, and so this backbone is used for all name reconciliation steps as
outlined below.

The European metaweb represents the knowledge we want to learn and
transfer; the phylogenetic similarity of mammals here represents the
support for transfer. We used the mammalian consensus supertree by
Upham, Esselstyn, and Jetz (2019), for which all approximatively 6000
names have been similarly matched to their GBIF valid names. This step
allows us to place each node of the mammalian European metaweb in the
phylogeny.

The destination problem to which we want to transfer knowledge is the
trophic interactions between mammals in Canada. We obtained the list of
extant species from the IUCN checklist, and selected the terrestrial and
semi-aquatic species (this corresponds to the same selection that was
applied by Maiorano et al. (2020b) in the European metaweb). The IUCN
names were, as previously, reconciled against GBIF to have an exact
match to the taxonomy.

After taxonomic cleaning and reconciliation as outlined in the following
sections, the mammalian European metaweb had 260 species, and the
Canadian species pool has 163; of these, 17 (about 4\% of the total) are
shared, and 89 species from Canada (54\%) had at least one congeneric
species in Europe. The similarity for both species pool predictably
increases with higher taxonomic order, with 19\% of shared genera, 47\%
of shared families, and 75\% of shared orders; for the last point,
Canada and Europe each had a single unique order (\emph{Didelphimorphia}
for Canada, \emph{Erinaceomorpha} for Europe).

@@

Specifically, our approach can be summarized as follows: from the known
network in Europe, we use a truncated Singular Value Decomposition
(t-SVD; Halko, Martinsson, and Tropp 2011) to generate latent traits
representing a low-dimensional embedding of the network; these traits
give an unbiased estimate of the node's position in the latent feature
spaces. Then, we map these latent traits onto a reference phylogeny
(other distance-based measures of species proximity that allow for the
inference of features in the latent space can be used, for example the
dissimilarity in functional traits). Based on the reconstructed latent
traits for species in the destination species pool, a Random Dot Product
Graph model (hereafter RDPG; S. J. Young and Scheinerman 2007) predicts
the interaction between species through a function of the nodes'
features through matrix multiplication. Thus, from latent traits and
nodes position, we can infer interactions.

@@

The entire pipeline is implemented in \emph{Julia} 1.6 (Bezanson et al.
2017) and is available under the permissive MIT License at
\href{https://osf.io/2zwqm/}{\texttt{https://osf.io/2zwqm/}}. The
taxonomic cleanup steps are done using \texttt{GBIF.jl} (Dansereau and
Poisot 2021). The network embedding and analysis is done using
\texttt{EcologicalNetworks.jl} (Banville, Vissault, and Poisot 2021;
Poisot et al. 2019). The phylogenetic simulations are done using
\texttt{PhyloNetworks.jl} (Solís-Lemus, Bastide, and Ané 2017) and
\texttt{Phylo.jl} (Reeve et al. 2016). A complete \texttt{Project.toml}
file specifying the full tree of dependencies is available alongside the
code. This material also includes a fully annotated copy of the entire
code required to run this project (describing both the intent of the
code and discussing some technical implementation details), a vignette
for every step of the process, and a series of Jupyter notebooks with
the text and code. The pipeline can be executed on a laptop in a matter
of minutes, and therefore does not require extensive computational
power.

@@

The first step in transfer learning is to learn the structure of the
original dataset. In order to do so, we rely on an approach inspired
from representational learning, where we learn a \emph{representation}
of the metaweb (in the form of the latent subspaces), rather than a list
of interactions (species \emph{a} eats \emph{b}). This approach is
conceptually different from other metaweb-scale predictions
(\emph{.e.g.} Albouy et al. 2019), in that the metaweb representation is
easily transferable. Specifically, we use RDPG to create a number of
latent variables that can be combined into an approximation of the
network adjacency matrix. RDPG results are known to have strong
phylogenetic signal, and to capture the evolutionary backbone of food
webs (Dalla Riva and Stouffer 2016). In addition, recent advances show
that the latent variables produced this way can be used to predict
\emph{de novo} network edges (Runghen, Stouffer, and Dalla Riva 2021).

The latent variables are created by performing a truncated Singular
Value Decomposition (t-SVD) on the adjacency matrix. SVD is an
appropriate embedding of ecological networks, which has recently been
shown to both capture their complex, emerging properties (Strydom, Dalla
Riva, and Poisot 2021) and to allow highly accurate prediction of the
interactions within a single network (Poisot, Ouellet, et al. 2021).
Under SVD, an adjacency matrix \(\mathbf{A}\) (where
\(\mathbf{A}_{m,n}\in\mathbb{B}\) where 1 indicates predation and 0 an
absence thereof) is decomposed into three components resulting in
\(\mathbf{A} = \mathbf{L}\mathbf{\Sigma}\mathbf{R}.\) Here,
\(\mathbf{\Sigma}\) is a \(m \times n\) diagonal matrix and contains
only singular (\(\sigma\)) values along its diagonal, \(\mathbf{L}\) is
a \(m \times m\) unitary matrix, and \(\mathbf{R}'\) a \(n \times n\)
unitary matrix. Truncating the SVD removes additional noise in the
dataset by omitting non-zero and/or smaller \(\sigma\) values from
\(\mathbf{\Sigma}\) using the rank of the matrix. Under a t-SVD
\(\mathbf{A}_{m,n}\) is decomposed so that \(\mathbf{\Sigma}\) is a
square \(r \times r\) diagonal matrix (where \(r\) is the rank of
\(\mathbf{A}\)) containing only non-zero \(\sigma\) values.
Additionally, \(\mathbf{L}\) is now a \(m \times r\) semi unitary matrix
and \(\mathbf{R}'\) a \(n \times r\) semi-unitary matrix.

The specific rank at which the SVD ought to be truncated is a difficult
question. The purpose of SVD is to remove the noise (expressed at high
dimensions) and to focus on the signal, (expressed at low dimensions).
In datasets with a clear signal/noise demarcation, a scree plot of
\(\mathbf{\Sigma}\) can show a sharp drop at the rank where noise starts
(Zhu and Ghodsi 2006). Because the European metaweb is almost entirely
known, the amount of noise is low; this is reflected in
fig.~\ref{fig:scree} (left), where the scree plot shows no important
drop, and in fig.~\ref{fig:scree} (right) where the proportion of
variance explained increases smoothly at higher dimensions. For this
reason, we default back to an arbitrary threshold that explains 60\% of
the variance in the underlying data, corresponding to 12 dimensions.

@@

\begin{figure}
\hypertarget{fig:scree}{%
\centering
\includegraphics{figures/figure-screeplot.png}
\caption{Left: representation of the screeplot of the singular values
from the t-SVD on the European metaweb. The screeplot shows no obvious
drop in the singular values that may be leveraged to automatically
detect a minimal dimension for embedding, after \emph{e.g.} Zhu and
Ghodsi (2006). Right: cumulative fraction of variance explained by each
dimension up to the rank of the European metaweb. The grey lines
represent cutoffs at 50, 60\ldots{} 90\% of variance explained. For the
rest of the analysis, we reverted to an arbitrary threshold of 60\% of
variance explained, which represented a good tradeoff between accuracy
and reduced number of features.}\label{fig:scree}
}
\end{figure}

Because RDPG relies on matrix multiplication, the higher dimensions
essentially serve to make specific interactions converge towards 0 or 1;
therefore, for reasonably low ranks, there is no guarantee that the
values in the reconstructed network will be within the unit range. In
order to determine what constitutes an appropriate threshold for
probability, we performed the RDPG approach on the European metaweb, and
evaluated the probability threshold by treating this as a binary
classification problem, specifically assuming that both 0 and 1 in the
European metaweb are all true. Given the methodological details given in
Maiorano et al. (2020b) and O'Connor et al. (2020), this seems like a
reasonable assumption, although one that does not hold for all metawebs.
We used the thresholding approach presented in Poisot, Ouellet, et al.
(2021), and picked a cutoff that maximized Youden's \(J\) statistic
(Youden (1950); a measure of the informedness (trust) of predictions);
the resulting cutoff was 0.22, and gave an accuracy above 0.99.

@@

Once the probabilistic metaweb for Canada has been produced, we followed
a number of data inflation steps to finalize it.

>>>>

\usepackage{amsmath,amssymb,amsfonts}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{stix2}
\usepackage[scaled]{helvet}
\usepackage[scaled]{inconsolata}

@@

\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[3] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
  % set entry spacing
  \ifnum #2 > 0
  \setlength{\parskip}{#2\baselineskip}
  \fi
 }%
 {}
\usepackage{calc} % for \widthof, \maxof
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\maxof{\widthof{#1}}{\csllabelwidth}}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth}{#1}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}\geometry{verbose,letterpaper,tmargin=2.2cm,bmargin=2.2cm,lmargin=2.2cm,rmargin=2.2cm}

@@

\tolerance=1
\emergencystretch=\maxdimen
\hyphenpenalty=10000
\hbadness=10000

@@

\fancypagestyle{normal}
{
  \fancyhf{}
  \fancyfoot[R]{\footnotesize\sffamily\thepage\ of \pageref*{LastPage}}
}
\begin{document}
\raggedright
\thispagestyle{empty}
{\Large\bfseries\sffamily Food web reconstruction through phylogenetic
transfer of low-rank network representation}
\vskip 5em

%
\href{https://orcid.org/0000-0001-6067-1349}{Tanya\,Strydom}%
%
\,\textsuperscript{1,2,‡}\quad %
\href{https://orcid.org/0000-0003-0193-5441}{Salomé\,Bouskila}%
%
\,\textsuperscript{1,‡}\quad %
\href{https://orcid.org/0000-0001-9051-0597}{Francis\,Banville}%
%
\,\textsuperscript{1,3,2}\quad %
\href{https://orcid.org/0000-0003-4036-977X}{Ceres\,Barros}%
%
\,\textsuperscript{4}\quad %
\href{https://orcid.org/0000-0002-2151-6693}{Dominique\,Caron}%
%
\,\textsuperscript{5,2}\quad %
\href{https://orcid.org/0000-0003-0452-6993}{Maxwell J\,Farrell}%
%
\,\textsuperscript{6}\quad %
\href{https://orcid.org/0000-0002-9935-1366}{Marie-Josée\,Fortin}%
%
\,\textsuperscript{6}\quad %
\href{https://orcid.org/0000-0003-3220-6161}{Victoria\,Hemming}%
%
\,\textsuperscript{4}\quad %
\href{https://orcid.org/0000-0002-4104-9463}{Benjamin\,Mercier}%
%
\,\textsuperscript{3,2}\quad %
\href{https://orcid.org/0000-0002-6004-4027}{Laura J.\,Pollock}%
%
\,\textsuperscript{5,2}\quad %
\href{https://orcid.org/0000-0001-5785-8321}{Rogini\,Runghen}%
%
\,\textsuperscript{7}\quad %
\href{https://orcid.org/0000-0002-3454-0633}{Giulio V.\,Dalla Riva}%
%
\,\textsuperscript{8}\quad %
\href{https://orcid.org/0000-0002-0735-5184}{Timothée\,Poisot}%
%
\,\textsuperscript{1,2}

@@

\vfill
Despite their importance in many ecological processes, collecting data
and information on ecological interactions, and therefore species
interaction networks, is an exceedingly challenging task. For this
reason, large parts of the world have a data deficit when it comes to
species interactions, and how the resulting networks are structured. As
data collection alone is unlikely to be sufficient at filling these
global gaps, community ecologists must adopt predictive methods. In this
contribution, we develop such a method, relying on graph embedding (the
extraction of explanatory latent variables from known graph structures)
and transfer learning (the application of previous solutions to novel
problems with limited predictors overlap) in order to assemble a
predicted list of trophic interactions between mammals of Canada. This
interaction list is derived from extensive knowledge of the mammalian
food web of Europe, despite the fact that there are fewer than 5\% of
common species between the two locations. The results of the predictive
model are compared against databases of recorded pairwise interactions,
showing that we correctly recover over 95\% of known interactions. We
provide guidance on how this method can be adapted by substituting some
approaches or predictors in order to make it more generally applicable.

\vfill

@@

There are two core challenges we are faced with in furthering our
understanding of ecological networks across space, particularly at
macro-ecologically relevant scales (\emph{e.g.} Trøjelsgaard \& Olesen
2016). First, networks within a location are difficult to sample
properly (Jordano 2016a, b), resulting in a widespread ``Eltonian
shortfall'' (Hortal \emph{et al.} 2015), \emph{i.e.} a lack of knowledge
about inter and intra specific relationships. This first challenge has
been, in large part, addressed by the recent emergence of a suite of
methods aiming to predict interactions within \emph{existing} networks,
many of which are reviewed in Strydom \emph{et al.} (2021a). Second,
recent analyses based on collected data (Poisot \emph{et al.} 2021a) or
metadata (Cameron \emph{et al.} 2019) highlight that ecological networks
are currently studied in a biased subset of space and bioclimates, which
impedes our ability to generalize any local understanding of network
structure. Meaning that, although the framework to address
incompleteness \emph{within} networks exists, there would still be
regions for which, due to a \emph{lack} of local interaction data, we
are unable to infer potential species interactions. Having a general
solution for inferring a \emph{plausible} metaweb (despite the
unavailability of interaction data) could be the catalyst for
significant breakthroughs in our ability to start thinking about species
interactions networks over large spatial scales.

Here, we present a general method for the transfer learning of network
representations, relying on the similarities of species in a
biologically/ecologically relevant proxy space (\emph{e.g.} shared
morphology or ancestry). Transfer learning is a machine learning
methodology that uses the knowledge gained from solving one problem and
applying it to a related (destination) problem (Pan \& Yang 2010; Torrey
\& Shavlik 2010). In this instance, we solve the problem of predicting
trophic interactions between species, based on knowledge extracted from
another species pool for which interactions are known by using
phylogenetic structure as a medium for transfer. This allows us to
construct a \emph{probabilistic} metaweb for a community for which we
have \emph{no} prior trophic interaction data for the desired species
pool. Our methodology is outlined in fig.~\ref{fig:concept}, where we
provide an illustration based on learning the embedding of a metaweb of
trophic interactions for European mammals (known interactions; Maiorano
\emph{et al.} 2020b, a) and, based on phylogenetic relationships between
mammals globally (\emph{i.e.}, phylogenetic tree Upham \emph{et al.}
2019), infer a metaweb for the Canadian mammalian species pool
(interactions are treated as unknown in this instance).

@@

There is a plurality of measures of species similarities that can be
used for metaweb reconstruction (see \emph{e.g.} Morales-Castilla
\emph{et al.} 2015); however, phylogenetic proximity has several
desirable properties when working at large scales. Gerhold \emph{et al.}
(2015) made the point that phylogenetic signal captures diversification
of characters (large macro-evolutionary process), but not necessarily
community assembly (fine ecological process); Dormann \emph{et al.}
(2010) previously found very similar conclusions. Interactions tend
reflect a phylogenetic signal because they have a conserved pattern of
evolutionary convergence that encompasses a wide range of ecological and
evolutionary mechanisms (Cavender-Bares \emph{et al.} 2009; Mouquet
\emph{et al.} 2012), and - most importantly - retain this signal even
when it is not detectable at the community scale (Hutchinson \emph{et
al.} 2017; Poisot \& Stouffer 2018). Finally, species interactions at
macro-ecological scales seem to respond mostly to macro-evolutionary
processes (Price 2003); which is evidenced by the presence of conserved
backbones in food webs (Dalla Riva \& Stouffer 2016), strong
evolutionary signature on prey choice (Stouffer \emph{et al.} 2012), and
strong phylogenetic signature in food web intervality (Eklöf \& Stouffer
2016). Phylogenetic reconstruction has also previously been used within
the context of ecological networks, namely understanding ancestral
plant-insect interactions (Braga \emph{et al.} 2021). Taken together,
these considerations suggest that phylogenies can reliably be used to
transfer knowledge on species interactions.

Our case study shows that phylogenetic transfer learning is indeed an
effective approach to predict the Canadian mammalian metaweb. This
showcases that although the components (species) that make up the
Canadian and European communities may be \emph{minimally} shared, if the
medium (proxy space) selected in the transfer step is biologically
plausible, we can still effectively learn from the known network and
make biologically relevant predictions of interactions. It should be
reiterated that the framework presented in fig.~\ref{fig:concept} is
amenable to changes; notably, the measure of similarity may not be
phylogeny, and can be replaced by information on foraging (Beckerman
\emph{et al.} 2006), cell-level mechanisms (Boeckaerts \emph{et al.}
2021), or a combination of traits and phylogenetic structure (Stock
2021).

@@

We use data from the European metaweb assembled by Maiorano \emph{et
al.} (2020b), following the definition of the metaweb first introduced
by Dunne (2006), \emph{i.e.} an inventory of all possible interactions
within species from a spatially delimited pool. Notably the metaweb is
not a prediction of the food web at any specific locale within the
frontiers of the species pool -- in fact, these local food webs are
expected to have a subset of both the species and the interactions of
their metaweb (Poisot \emph{et al.} 2012). This being said, as the
metaweb represents the total of functional, phylogenetic, and
macroecological processes (Morales-Castilla \emph{et al.} 2015), it is
thus still worthy of ecological attention. We deduced the subgraph
corresponding to all mammals by matching species names in the original
network to the GBIF taxonomic backbone (GBIF Secretariat 2021) and
retaining all those who matched to mammals. This serves a dual purpose
1) to extract only mammals from the European network and 2) to match and
standardize species names when aggregating the different data sources
further downstream (which is an important consideration when combining
datasets Grenié \emph{et al.} (2021)). All nodes had valid matches to
GBIF at this step, and so this backbone is used for all name
reconciliation steps as outlined below.

The European metaweb represents the knowledge we want to learn and
transfer; the phylogenetic similarity of mammals here represents the
information for transfer. We used the mammalian consensus supertree by
Upham \emph{et al.} (2019), for which all approximatively 6000 names
have been similarly matched to their GBIF valid names. This step allows
us to place each node of the mammalian European metaweb in the
phylogeny.

The destination problem to which we want to transfer knowledge is the
trophic interactions between mammals in Canada. We obtained the list of
extant species from the IUCN checklist, and selected the terrestrial and
semi-aquatic species (this corresponds to the same selection that was
applied by Maiorano \emph{et al.} (2020b) in the European metaweb). The
IUCN names were, as previously, reconciled against GBIF to have an exact
match to the taxonomy.

After taxonomic cleaning and reconciliation as outlined in the following
sections, the mammalian European metaweb has 260 species, and the
Canadian species pool has 163; of these, 17 (about 4\% of the total) are
shared, and 89 species from Canada (54\%) had at least one congeneric
species in Europe. The similarity for both species pools predictably
increases with higher taxonomic order, with 19\% of shared genera, 47\%
of shared families, and 75\% of shared orders; for the last point,
Canada and Europe each had a single unique order (\emph{Didelphimorphia}
for Canada, \emph{Erinaceomorpha} for Europe).

@@

Specifically, our approach can be summarized as follows: from the known
network in Europe, we use a truncated Singular Value Decomposition
(t-SVD; Halko \emph{et al.} 2011) to generate latent traits representing
a low-dimensional embedding of the network; these traits give an
unbiased estimate of the node's position in the latent feature spaces.
Then, we map these latent traits onto a reference phylogeny (other
distance-based measures of species proximity that allow for the
inference of features in the latent space can be used, for example the
dissimilarity in functional traits). Based on the reconstructed latent
traits for species in the destination species pool, a Random Dot Product
Graph model (hereafter RDPG; Young \& Scheinerman 2007) predicts the
interaction between species through a function of the nodes' features
through matrix multiplication. Thus, from latent traits and node
position, we can infer interactions.

@@

The entire pipeline is implemented in \emph{Julia} 1.6 (Bezanson
\emph{et al.} 2017) and is available under the permissive MIT License at
\href{https://osf.io/2zwqm/}{\texttt{https://osf.io/2zwqm/}}. The
taxonomic cleanup steps are done using \texttt{GBIF.jl} (Dansereau \&
Poisot 2021). The network embedding and analysis is done using
\texttt{EcologicalNetworks.jl} (Poisot \emph{et al.} 2019; Banville
\emph{et al.} 2021). The phylogenetic simulations are done using
\texttt{PhyloNetworks.jl} (Solís-Lemus \emph{et al.} 2017) and
\texttt{Phylo.jl} (Reeve \emph{et al.} 2016). A complete
\texttt{Project.toml} file specifying the full tree of dependencies is
available alongside the code. This material also includes a fully
annotated copy of the entire code required to run this project
(describing both the intent of the code and discussing some technical
implementation details), a vignette for every step of the process, and a
series of Jupyter notebooks with the text and code. The pipeline can be
executed on a laptop in a matter of minutes, and therefore does not
require extensive computational power.

@@

The first step in transfer learning is to learn the structure of the
original dataset. In order to do so, we rely on an approach inspired
from representational learning, where we learn a \emph{representation}
of the metaweb (in the form of the latent subspaces), rather than a list
of interactions (species \emph{a} eats \emph{b}). This approach is
conceptually different from other metaweb-scale predictions (\emph{e.g.}
Albouy \emph{et al.} 2019), in that the metaweb representation is easily
transferable. Specifically, we use RDPG to create a number of latent
variables that can be combined into an approximation of the network
adjacency matrix. RDPG results are known to have strong phylogenetic
signal, and to capture the evolutionary backbone of food webs (Dalla
Riva \& Stouffer 2016). In addition, recent advances show that the
latent variables produced this way can be used to predict \emph{de novo}
network edges (\emph{i.e.} interactions; Runghen \emph{et al.} 2021).

The latent variables are created by performing a truncated Singular
Value Decomposition (t-SVD) on the adjacency matrix. SVD is an
appropriate embedding of ecological networks, which has recently been
shown to both capture their complex, emerging properties (Strydom
\emph{et al.} 2021b) and to allow highly accurate prediction of the
interactions within a single network (Poisot \emph{et al.} 2021b). Under
SVD, an adjacency matrix \(\mathbf{A}\) (where
\(\mathbf{A}_{m,n}\in\mathbb{B}\) where 1 indicates predation and 0 an
absence thereof) is decomposed into three components resulting in
\(\mathbf{A} = \mathbf{L}\mathbf{\Sigma}\mathbf{R'}.\) Here,
\(\mathbf{\Sigma}\) is a \(m \times n\) diagonal matrix and contains
only singular (\(\sigma\)) values along its diagonal, \(\mathbf{L}\) is
a \(m \times m\) unitary matrix, and \(\mathbf{R}'\) a \(n \times n\)
unitary matrix. Truncating the SVD removes additional noise in the
dataset by omitting non-zero and/or smaller \(\sigma\) values from
\(\mathbf{\Sigma}\) using the rank of the matrix. Under a t-SVD
\(\mathbf{A}_{m,n}\) is decomposed so that \(\mathbf{\Sigma}\) is a
square \(r \times r\) diagonal matrix (whith \(1 \le r \le r_{full}\)
where \(r_{full}\) is the full rank of \(\mathbf{A}\) and \(r\) the rank
at which we truncate the matrix) containing only non-zero \(\sigma\)
values. Additionally, \(\mathbf{L}\) is now a \(m \times r\) semi
unitary matrix and \(\mathbf{R}'\) a \(n \times r\) semi-unitary matrix.

The specific rank at which the SVD ought to be truncated is a difficult
question. The purpose of SVD is to remove the noise (expressed at high
dimensions) and to focus on the signal, (expressed at low dimensions).
In datasets with a clear signal/noise demarcation, a scree plot of
\(\mathbf{\Sigma}\) can show a sharp drop at the rank where noise starts
(Zhu \& Ghodsi 2006). Because the European metaweb is almost entirely
known, the amount of noise (uncertainty) is low; this is reflected in
fig.~\ref{fig:scree} (left), where the scree plot shows no important
drop, and in fig.~\ref{fig:scree} (right) where the proportion of
variance explained increases smoothly at higher dimensions. For this
reason, we default back to a threshold that explains 60\% of the
variance in the underlying data, corresponding to 12 dimensions -
\emph{i.e.} a tradeoff between accuracy and a reduced number of
features.

@@

\begin{figure}
\hypertarget{fig:scree}{%
\centering
\includegraphics{figures/figure-screeplot.png}
\caption{Left: representation of the screeplot of the singular values
from the t-SVD on the European metaweb. The screeplot shows no obvious
drop in the singular values that may be leveraged to automatically
detect a minimal dimension for embedding, after \emph{e.g.} Zhu \&
Ghodsi (2006). Right: cumulative fraction of variance explained by each
dimension up to the rank of the European metaweb. The grey lines
represent cutoffs at 50, 60\ldots{} 90\% of variance explained. For the
rest of the analysis, we reverted to an arbitrary threshold of 60\% of
variance explained, which represented a good tradeoff between accuracy
and reduced number of features.}\label{fig:scree}
}
\end{figure}

Because RDPG relies on matrix multiplication, the higher dimensions
essentially serve to make specific interactions converge towards 0 or 1;
therefore, for reasonably low ranks, there is no guarantee that the
values in the reconstructed network will be within the unit range. In
order to determine what constitutes an appropriate threshold for
probability, we performed the RDPG approach on the European metaweb, and
evaluated the probability threshold by treating this as a binary
classification problem, specifically assuming that both 0 and 1 in the
European metaweb are all true. Given the methodological details given in
Maiorano \emph{et al.} (2020b) and O'Connor \emph{et al.} (2020), this
seems like a reasonable assumption, although one that does not hold for
all metawebs. We used the thresholding approach presented in Poisot
\emph{et al.} (2021b), and picked a cutoff that maximized Youden's \(J\)
statistic (a measure of the informedness (trust) of predictions; Youden
(1950)); the resulting cutoff was 0.22, and gave an accuracy above 0.99.

@@

Once the probabilistic metaweb for Canada has been produced, we followed
a number of data inflation steps to finalize it. This step is external
to the actual transfer learning framework but rather serves as a way to
augment and validate the predicted metaweb.
